---
title: "CMIPB_2023"
author: "Anna Konstorum"
date: "2023-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries}
library(dplyr)
library(tidyr)
library(purrr)
library(nipalsMCIA)
library(glmnet)
library(readr)
#library(formattable)
```

# Many of the functions in the source file and in this script
# have been modified from the templates provided by the CMIPB team

```{r load custom functions}
source("CMIPB_2023_functions.R")
```

# Prepare dates/variables of interest
```{r variables}
DATASET <- c("2020_dataset", "2021_dataset")
TIMEPOINTS <- c(0, 1, 3, 14)
#dataFile <- "combined_dataset2020_2021.csv"

META_COLS <- c("specimen_id", "subject_id", "timepoint", "dataset", 
               "biological_sex", "infancy_vac", "age_at_boost")
ABTITER_COLS <- c("IgG_PT")
RNA_COLS <- c("CCL3")
CELL_COLS <- c("Monocytes")

DEMOGRAPHY_COLS <- c("age_at_boost", "biological_sex", "infancy_vac")

TASK_COLS <- c("Monocytes_D1", "CCL3_D3", "IgG_PT_D14")
TASKS_BASELINES <- c("Monocytes_D0", "CCL3_D0", "IgG_PT_D0")
BASE_COLS <- c("Monocytes_D0", "IgG_PT_D0", "CCL3_D0")

LOG_TRANS_COLS <- c("CCL3_D0", "IgG_PT_D0",
                    "CCL3_D3",  "IgG_PT_D14")
output_request = list("DATASET" = DATASET, "TIMEPOINTS" = TIMEPOINTS, "META_COLS" = META_COLS,
                   "ABTITER_COLS" = ABTITER_COLS, "RNA_COLS" = RNA_COLS, "CELL_COLS" = CELL_COLS,
                   "DEMOGRAPHCY_COLS" = DEMOGRAPHY_COLS, "TASK_COLS" = TASK_COLS,
                   "TASKS_BASELINES" = TASKS_BASELINES, "BASE_COLS" = BASE_COLS, 
                   "LOG_TRANS_COLS" = LOG_TRANS_COLS)
```

```{r load training data}
df_pr <- readRDS("master_processed_training_data.RDS")
```

## Extract data associated with outcome variables from training data
```{r combine_experiments}
df_pr_all <- combine_exps(df_pr, output_request)
df_pr_all <- fold_change(df_pr_all) 
```

## Prepare training data for model
```{r prepare_training_data}

# Filter metadata to baseline data
metaDf <- data.frame(df_pr[["subject_specimen"]])
metaDf["age_at_boost"] <- as.numeric(round(difftime(metaDf$date_of_boost, metaDf$year_of_birth,units="weeks")/52, 2))
metaDf <- metaDf[, output_request$META_COLS] %>%
          data.frame()
# meta_baseline
metaDf_base <- metaDf[metaDf['timepoint']==0,]

# For this submission, decided to only use 2021 data for training
# Should test with the batch-corrected 2020 + 2021 data
metaDf_base <- metaDf_base[metaDf_base['dataset']=='2021_dataset',]
metaDf_test <- metaDf[metaDf$timepoint %in% c(0,1,3,14),]
```

```{r}
# Extract associated speciman_ids from each dataset
abtiterDf <- df_pr[["abtiter_wide"]]$normalized_data %>%
    t() %>%
    data.frame() 
rnaDf <- df_pr[["pbmc_gene_expression"]]$raw_data %>%
    t() %>%
    data.frame() 
cellDf <- df_pr[["pbmc_cell_frequency"]]$normalized_data %>%
    t() %>%
    data.frame()
cytDf <- df_pr[["plasma_cytokine_concentrations"]]$normalized_data %>%
    t() %>%
    data.frame()
```

```{r, find_intersect}
# Find subjects with all four omics
# This step can be made more efficient
metaDf_base_cyt<-metaDf_base[metaDf_base$specimen_id %in% row.names(cytDf),]
metaDf_base_12<-metaDf_base_cyt[metaDf_base_cyt$specimen_id %in% row.names(abtiterDf),]
metaDf_base_123<-metaDf_base_12[metaDf_base_12$specimen_id %in% row.names(rnaDf),]
metaDf_base_1234<-metaDf_base_123[metaDf_base_123$specimen_id %in% row.names(cellDf),]

keep<-metaDf_base_1234[['specimen_id']]

# Subset each omics to baseline data for subjects in intersection and save as list
cytDf_base<-subset(cytDf, subset=rownames(cytDf) %in% keep)
abtiterDf_base<-subset(abtiterDf, subset=rownames(abtiterDf) %in% keep)
rnaDf_base<-subset(rnaDf, subset=rownames(rnaDf) %in% keep)
cellDf_base<-subset(cellDf, subset=rownames(cellDf) %in% keep)

rownames(abtiterDf_base)<-rownames(cytDf_base)<-rownames(rnaDf_base)<-rownames(cellDf_base)<-metaDf_base_1234[['subject_id']]

# After a few cross-validation experiments, found good performance using these three omics
# Can test with including back cellDf
base_data<-list(cyt=cytDf_base,ab=abtiterDf_base, seq=rnaDf_base) # cell=cellDf_base
```

# Run nipalsMCIA
```{r}

# Used r=10 factors, can add more factors in future experiments
base_data_mae <- simple_mae(base_data,row_format="sample")
set.seed(42)
mcia_base <- nipals_multiblock(base_data_mae,
                               col_preproc_method = "colprofile",
                               num_PCs = 10, tol = 1e-12, plots='none')
```

### Add demographic and baseline features to MCIA factor model to create MCIAplus model (gs_full)
```{r}
gs_full_temp <- mcia_base@global_scores
df_pr_all_prep<-subset(df_pr_all,subset=rownames(df_pr_all) %in% rownames(gs_full_temp))
df_pr_all_prep_base<-df_pr_all_prep[,c("biological_sex", "infancy_vac", "age_at_boost",
                                "IgG_PT_D0","Monocytes_D0","CCL3_D0")]
df_pr_all_prep_base['biological_sex'] <- as.numeric(as.factor(df_pr_all_prep_base$biological_sex))-1
df_pr_all_prep_base['infancy_vac'] <- as.numeric(as.factor(df_pr_all_prep_base$infancy_vac))-1


gs_full <- cbind(gs_full_temp,df_pr_all_prep_base)
train_tasks<-df_pr_all_prep[,c("IgG_PT_D14","IgG_PT_D14_FC",
                               "Monocytes_D1", "Monocytes_D1_FC",
                               "CCL3_D3", "CCL3_D3_FC")]
```

### Test model quality

```{r}

gs_full_filt<-na.omit(gs_full)

# Full model
gs_full_filt2<-gs_full_filt
pred_cor_full<-pred_cor_calc(gs_full_filt2,train_tasks)

# MCIA model
gs_full_filt2<-gs_full_filt[,c("1","2","3","4","5","6","7","8","9","10")]
pred_cor_mcia <- pred_cor_calc(gs_full_filt2,train_tasks)

# Day0/demographic model
gs_full_filt2<-gs_full_filt[,c('biological_sex','infancy_vac','age_at_boost','IgG_PT_D0','Monocytes_D0','CCL3_D0')]
pred_cor_clin <- pred_cor_calc(gs_full_filt2,train_tasks)

colnames(pred_cor_full) <- c("full")
colnames(pred_cor_mcia)<- c("MCIA")
colnames(pred_cor_clin)<- c("clin")

pred_cor_all<-cbind(pred_cor_full,pred_cor_mcia,pred_cor_clin)
pred_cor_all <- pred_cor_all %>%
  mutate(max = pmax(full, MCIA, clin, na.rm = TRUE))

# Decide which model to use for the predictions based on maximum correlation value
pred_cor_all$use<-c("clin","clin","full","none","clin","full")

```

### For each model, can assess which features contribute to non-zero coefficients
```{r}
# Using clinical model
gs_full_filt2<-gs_full_filt[,c('biological_sex','infancy_vac','age_at_boost','IgG_PT_D0','Monocytes_D0','CCL3_D0')]
train_tasks_clin<-train_tasks[,c("IgG_PT_D14", "IgG_PT_D14_FC","CCL3_D3")]
top_feat_clin<-top_feat_model(gs_full_filt2,train_tasks_clin)

# Using full model
gs_full_filt2<-gs_full_filt
train_tasks_full<-train_tasks[,c("Monocytes_D1","CCL3_D3_FC")]
top_feat_full<-top_feat_model(gs_full_filt2, train_tasks_full)


```

```{r load testing data}
df_pr_test <- readRDS("master_processed_prediction_data.RDS")
```

```{r prepare_test_data}

# Filter metadata
metaDf_test <- data.frame(df_pr_test[["subject_specimen"]])
metaDf_test["age_at_boost"] <- as.numeric(round(difftime(metaDf_test$date_of_boost, metaDf_test$year_of_birth,units="weeks")/52, 2))
metaDf_test <- metaDf_test[, META_COLS] %>%
          data.frame()
# meta_baseline
metaDf_test_base <- metaDf_test[metaDf_test['timepoint']==0,]
metaDf_test_base <- metaDf_test_base[metaDf_test_base['dataset']=='2022_dataset',]
metaDf_test_base <- metaDf_test_base[metaDf_test_base$timepoint %in% c(0,1,3,14),]
```

```{r}
# Extract associated specimen ids from each dataset

abtiterDf_test <- df_pr_test[["abtiter"]]$processed_similar_to_training %>%
    t() %>%
    data.frame() 
rnaDf_test <- df_pr_test[["pbmc_gene_expression"]]$processed_similar_to_training %>%
    t() %>%
    data.frame() 
cellDf_test <- df_pr_test[["pbmc_cell_frequency"]]$processed_similar_to_training %>%
    t() %>%
    data.frame()
cytDf_test <- df_pr_test[["plasma_cytokine_concentrations"]]$processed_similar_to_training %>%
    t() %>%
    data.frame()
```

```{r, find_intersect}
# Find intersection for specimen ids containing all omics and combine the omics data into a list
metaDf_test_base_cyt<-metaDf_test_base[metaDf_test_base$specimen_id %in% row.names(cytDf_test),]
metaDf_test_base_12<-metaDf_test_base_cyt[metaDf_test_base_cyt$specimen_id %in% row.names(abtiterDf_test),]
metaDf_test_base_123<-metaDf_test_base_12[metaDf_test_base_12$specimen_id %in% row.names(rnaDf_test),]
metaDf_test_base_1234<-metaDf_test_base_123[metaDf_test_base_123$specimen_id %in% row.names(cellDf_test),]

keep<-metaDf_test_base_1234[['specimen_id']]

cytDf_test_base<-subset(cytDf_test, subset=rownames(cytDf_test) %in% keep)
cytDf_test_base<-cytDf_test_base[colnames(cytDf_base)]
# Zero out na values
cytDf_test_base[is.na(cytDf_test_base)] <- 0

abtiterDf_test_base<-subset(abtiterDf_test, subset=rownames(abtiterDf_test) %in% keep)
abtiterDf_test_base<-abtiterDf_test_base[colnames(abtiterDf_base)]

rnaDf_test_base<-subset(rnaDf_test, subset=rownames(rnaDf_test) %in% keep)
rnaDf_test_base<-rnaDf_test_base[colnames(rnaDf_base)]

cellDf_test_base<-subset(cellDf_test, subset=rownames(cellDf_test) %in% keep)
cellDf_test_base<-cellDf_test_base[colnames(cellDf_base)]

rownames(abtiterDf_test_base)<-rownames(cytDf_test_base)<-rownames(rnaDf_test_base)<-rownames(cellDf_test_base)<-metaDf_test_base_1234[['subject_id']]

base_test_data<-list(cyt=cytDf_test_base,ab=abtiterDf_test_base, seq=rnaDf_test_base) # cell=cellDf_base
```

# Use MCIA model built on training data to build new model
```{r}
base_test_data_mae <- simple_mae(base_test_data,row_format="sample")
gs_full_new_temp <- predict_gs(mcia_results = mcia_base,
                               test_data = base_test_data_mae)
```

## Extract data associated with outcome variables from test data
```{r}
df_pr_test_all <- combine_exps_test(df_pr_test)
rownames(df_pr_test_all)<-df_pr_test_all$subject_id
```

```{r}
# Create test data structures in the same way to training to feed into models

# gs_test_full includes new MCIA scores as well as demographic/day0 factors
# Does not include data for subjects 97 and 105 as they were missing an omic
# If this omic imputed, can include in future
df_pr_test_all_prep<-subset(df_pr_test_all,subset=rownames(df_pr_test_all) %in% rownames(gs_full_new_temp))
df_pr_test_all_prep_base<-df_pr_test_all_prep[,c("biological_sex", "infancy_vac", "age_at_boost",
                                "IgG_PT_D0","Monocytes_D0","CCL3_D0")]
df_pr_test_all_prep_base['biological_sex'] <- as.numeric(as.factor(df_pr_test_all_prep_base$biological_sex))-1
df_pr_test_all_prep_base['infancy_vac'] <- as.numeric(as.factor(df_pr_test_all_prep_base$infancy_vac))-1
gs_test_full <- cbind(gs_full_new_temp,df_pr_test_all_prep_base)

# Data only from demographic/day0 variables, includes subjects 97 and 105
df_pr_test_all_prep_b<-df_pr_test_all
df_pr_test_all_prep_base_b<-df_pr_test_all_prep_b[,c("biological_sex", "infancy_vac", "age_at_boost",
                                "IgG_PT_D0","Monocytes_D0","CCL3_D0")]
df_pr_test_all_prep_base_b['biological_sex'] <- as.numeric(as.factor(df_pr_test_all_prep_base_b$biological_sex))-1
df_pr_test_all_prep_base_b['infancy_vac'] <- as.numeric(as.factor(df_pr_test_all_prep_base_b$infancy_vac))-1
gs_test_full_clin<-df_pr_test_all_prep_base_b

# Training data
gs_full_filt<-na.omit(gs_full)
gs_full_filt_clin<-gs_full_filt[,c('biological_sex','infancy_vac','age_at_boost','IgG_PT_D0','Monocytes_D0','CCL3_D0')]

```



```{r}
#gs_full_filt<-na.omit(gs_full)
#gs_full_filt_clin<-gs_full_filt[,c('biological_sex','infancy_vac','age_at_boost','IgG_PT_D0','Monocytes_D0','CCL3_D0')]

#gs_test_full_clin<-gs_test_full[,c('biological_sex','infancy_vac','age_at_boost','IgG_PT_D0','Monocytes_D0','CCL3_D0')]

# missing 97 and 105
#gs_test_full_clin<-df_pr_test_all_prep_base_b

```

```{r}
# Make predictions!
# pred_cor_calc_test takes in:
# (1) the take name
# (2) the training model, this is decided by the results in the pred_cor_all dataframe
# if the 'use' column indicates 'clin', then use the 'gs_full_filt_clin' training data, 
# and the 'gs_test_full_clin' test data
# if the 'use' column indicates 'full', then use the 'gs_full_filt' training data
# and the 'gs_test_full' test data; note this data is missing input for two subjects,
# so the predictions will be added in a later chunk

# 'use'=='clin'
task = "IgG_PT_D14"
IGG_D14<-pred_cor_calc_test(task,gs_full_filt_clin,gs_test_full_clin)

task = "IgG_PT_D14_FC"
IGG_FC<-pred_cor_calc_test(task,gs_full_filt_clin,gs_test_full_clin)

task = "CCL3_D3"
CCL3_D3<-pred_cor_calc_test(task,gs_full_filt_clin,gs_test_full_clin)

# 'use'== 'full'
task = "Monocytes_D1"
M_D1 <-pred_cor_calc_test(task,gs_full_filt,gs_test_full)

task = "CCL3_D3_FC"
CCL3_D3_FC <-pred_cor_calc_test(task,gs_full_filt,gs_test_full)
```

```{r}
# In order to get the rankings for the two missing subjects for Monocytes_D1 and 
# CCL3_D3_FC, we will use the rankings from the associated Day 0 values
targetX <- c("Monocytes_D0", "CCL3_D0")
df_pr_test_all_rank <- df_pr_test_all
df_pr_test_all_rank[, ranked_cols_base] <- apply(df_pr_test_all_rank[, targetX],2,rankingFunction)
```

# Start to combine results
```{r}
# Combine outputs of models only using day0/demographic data into one dataframe
# This dataframe contains all subjects
clin_all<-cbind(IGG_D14,IGG_FC,CCL3_D3)
colnames(clin_all)<-c("IgG-PT-D14-titer","IgG-PT-D14-FC","CCL3-D3")
clin_all_rank<-clin_all
ranked_cols<-paste(colnames(clin_all),"Rank",sep="_")
clin_all_rank[, ranked_cols] <- apply(clin_all_rank[, colnames(clin_all)],2,rankingFunction)
```

```{r}
# Combine output of full models into one dataframe
# This is missing rankings for two subjects
full_all<-cbind(M_D1, CCL3_D3_FC)
colnames(full_all)<-c("Monocytes-D1","CCL3-D3-FC")
full_all_rank<-full_all
ranked_cols<-paste(colnames(full_all),"Rank",sep="_")
full_all_rank[, ranked_cols] <- apply(full_all_rank[, colnames(full_all)],2,rankingFunction)
```

```{r}
submission_template = read_tsv("2ndChallengeSubmissionTemplate.tsv")

clin_all_rank$subject<-rownames(clin_all_rank)
full_all_rank$subject<-rownames(full_all_rank)
final_out<-merge(clin_all_rank,by="subject",full_all_rank,all=T)

# Add prediction rankings using day 0 values for the two subjects
# who did not have cytof but for the test variables where the model needed it
final_out[final_out$subject==105,"Monocytes-D1_Rank"] = df_pr_test_all_rank[df_pr_test_all_rank$subject_id==105,"Monocytes_D0_Rank"]

final_out[final_out$subject==105,"CCL3-D3-FC_Rank"] = df_pr_test_all_rank[df_pr_test_all_rank$subject_id==105,"CCL3_D0_Rank"]

final_out[final_out$subject==97,"Monocytes-D1_Rank"] = df_pr_test_all_rank[df_pr_test_all_rank$subject_id==97,"Monocytes_D0_Rank"]

final_out[final_out$subject==97,"CCL3-D3-FC_Rank"] = df_pr_test_all_rank[df_pr_test_all_rank$subject_id==97,"CCL3_D0_Rank"]

rownames(final_out)<-as.numeric(final_out$subject)

reord<-match(submission_template$`Subject ID`,rownames(final_out))
final_out<-final_out[reord,]
```

# Submit!
```{r}
submission_template$`1.1) IgG-PT-D14-titer-Rank`<-final_out$`IgG-PT-D14-titer_Rank`
submission_template$`1.2) IgG-PT-D14-FC-Rank`<-final_out$`IgG-PT-D14-FC_Rank`
submission_template$`2.1) Monocytes-D1-Rank`<-final_out$`Monocytes-D1_Rank`
submission_template$`3.1) CCL3-D3-Rank`<-final_out$`CCL3-D3_Rank`
submission_template$`3.2) CCL3-D3-FC-Rank`<-final_out$`CCL3-D3-FC_Rank`
```

```{r}
write.table(submission_template,'submission_output.tsv',quote=FALSE,sep="\t",
            row.names=FALSE, na="")
```
